---
title: "parse_opcodes"
output: pdf_document
---

```{r setup, include=FALSE}
suppressPackageStartupMessages({
  library(magrittr)
  library(tidyverse)
  library(rvest)
  library(httr)
  library(xml2)
})

options(max.print = 1000000)

knitr::opts_chunk$set(echo = TRUE)
```



```{r}
# h <- xml2::read_html('http://www.oxyron.de/html/opcodes02.html')
h <- xml2::read_html("opcodes.html")
```


Address types
------------------------------------------------------------------------------

imm = #$00
zp = $00
zpx = $00,X
zpy = $00,Y
izx = ($00,X)
izy = ($00),Y
abs = $0000
abx = $0000,X
aby = $0000,Y
ind = ($0000)
rel = $0000 (PC-relative) 



```{r}
mode_df <- readr::read_csv(
'mode, mode_desc    , mode_template       
imp  , implied      , 
imm  , immediate    ,"#$00"
zp   , zero page    ,"$00"
zpx  , zero page x  ,"$00,X"
zpy  , zero page y  ,"$00,Y"
izx  , indirect x   ,"($00,X)"
izy  , indirect y   ,"($00),Y"
abs  , absolute     ,"$0000"
abx  , absolute x   ,"$0000,X"
aby  , absolute y   ,"$0000,Y"
ind  , indirect     ,"($0000)"
rel  , relative     ,"$000o0"
')
```




Flags of the status register:
------------------------------------------------------------------------------

The processor status register has 8 bits, where 7 are used as flags:

N = negative flag (1 when result is negative)
V = overflow flag (1 on signed overflow)
# = unused (always 1)
B = break flag (1 when interupt was caused by a BRK)
D = decimal flag (1 when CPU in BCD mode)
I = IRQ flag (when 1, no interupts will occur (exceptions are IRQs forced by BRK and NMIs))
Z = zero flag (1 when all bits of a result are 0)
C = carry flag (1 on unsigned overflow)



```{r}
tables <- h %>% html_nodes('table')

# html_table(tables[1])
logical_and_arithmetic <- html_table(tables[2])[[1]] %>% set_names(.[1,]) %>% filter(row_number() != 1) %>% mutate(type='logical')
move                   <- html_table(tables[3])[[1]] %>% set_names(.[1,]) %>% filter(row_number() != 1) %>% mutate(type='move')
jump_flag              <- html_table(tables[4])[[1]] %>% set_names(.[1,]) %>% filter(row_number() != 1) %>% mutate(type='jump')
illegal_opcodes        <- html_table(tables[5])[[1]] %>% set_names(.[1,]) %>% filter(row_number() != 1) %>% mutate(type='illegal')


opcodes <- dplyr::bind_rows(logical_and_arithmetic, move, jump_flag, illegal_opcodes) %>%
  rename(
    op   = Opcode,
    desc = Function
  )


opcodes %<>% 
  gather(mode, hex, imp:rel) %>%
  filter(hex != '') %>%
  arrange(hex) %>% 
  left_join(mode_df, by='mode') %>%
  mutate(
    dec = strtoi(gsub("\\$", '0x', hex))
  ) %>%
  as.tbl

opcodes

```


```{r}
opcodes %<>% mutate(
  op = tolower(op)
) %>%
  rename(opmode = mode_desc)
```


```{r}
readr::write_delim(opcodes, "opcodes.tsv", delim="\t")

devtools::use_data(opcodes, overwrite=TRUE)
```


```{r}
ops      <- sort(unique(opcodes$op)) 
ops      <- stringr::str_sub(ops, end=3)
lowerops <- tolower(ops)

opcode_pattern <- paste0("\\b(", paste(c(ops, lowerops), collapse="|"), ")\\b")

opcode_pattern
# devtools::use_data(opcode_pattern, overwrite=TRUE)
```





