---
title: "Hello World - compilation details"
author: "mikefc@coolbutuseless.com"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Hello World - compilation details}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
suppressPackageStartupMessages({
  library(dplyr)
  library(purrr)
  library(c64asm)  
})

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


Write your c64 ASM code
------------------------------------------------------------------------------

* The syntax is roughly modelled on that of TASS64, but only for the fundamentals i.e.
    * `*` refers to current address location, and can be read from or written to
    * All programs should start with a `*=....` line to set the starting address
    * `.byte` values must be hexadecimal only, and single byte hexadecimal must **always**
      include 2 characters.  i.e. `$0e` will work, but `$e` won't work.
    * Similarly, for immediate values. i.e. `#$01` will work. `#$1` won't work.

```{r}
library(dplyr)
library(c64asm)

asm <- '
*=$0801
  .byte $0c, $08, $0a, $00, $9e, $20
  .byte $32, $30, $38, $30, $00, $00
  .byte $00

*=$0820
      ldx #$00
loop  lda message,x
      and #$3f
      sta $0400,x
      inx
      cpx #$0c
      bne loop

      rts

message
    .text "Hello World!"
'
```


c64asm defines a set of tokens in the ASM file
------------------------------------------------------------------------------

* `c64asm::asm_patterns` defines the set of patterns for lexing/tokenising an ASM file
* Things to note:
    * upper or lower-case opcodes are allowed, but not mixed-case
    * No decimal representation is supported, everything should be in hexadecimal
    * A bare `*` is interpreted as the program counter (i.e. current address)
* Custom R features
    * `.rbyte` directive. similar to the standard `.byte`, but the bytes come from the
      next token in the stream which must be a variable name which contains integers
      in the range 0-255.
    * `.rtext` directive. Similar to the standard `.text`, but the text comes from the 
      next token in the stream which must be a variable name which contains a string.
    * `{...}` represents text to be evaluated. Used for symbol arithmetic. e.g. `lda {message},1`

```{r}
#-----------------------------------------------------------------------------
# Regex patterns for parsing 6502 assembly
#-----------------------------------------------------------------------------
asm_patterns <- c(
  newline    = '\n',
  whitespace = '\\s+',
  PC         = '\\*',
  immediate  = '#\\$[0-9a-fA-F]{1,2}',
  word       = '\\$[0-9a-fA-F]{3,4}',
  byte       = '\\$[0-9a-fA-F]{1,2}',
  opcode     = "\\b(ADC|AHX|ALR|ANC|AND|ARR|ASL|AXS|BCC|BCS|BEQ|BIT|BMI|BNE|BPL|BRK|BVC|BVS|CLC|CLD|CLI|CLV|CMP|CPX|CPY|DCP|DEC|DEX|DEY|EOR|INC|INX|INY|ISC|JMP|JSR|LAS|LAX|LAX|LDA|LDX|LDY|LSR|NOP|ORA|PHA|PHP|PLA|PLP|RLA|ROL|ROR|RRA|RTI|RTS|SAX|SBC|SEC|SED|SEI|SHX|SHY|SLO|SRE|STA|STX|STY|TAS|TAX|TAY|TSX|TXA|TXS|TYA|XAA|adc|ahx|alr|anc|and|arr|asl|axs|bcc|bcs|beq|bit|bmi|bne|bpl|brk|bvc|bvs|clc|cld|cli|clv|cmp|cpx|cpy|dcp|dec|dex|dey|eor|inc|inx|iny|isc|jmp|jsr|las|lax|lax|lda|ldx|ldy|lsr|nop|ora|pha|php|pla|plp|rla|rol|ror|rra|rti|rts|sax|sbc|sec|sed|sei|shx|shy|slo|sre|sta|stx|sty|tas|tax|tay|tsx|txa|txs|tya|xaa)\\b",
  byte_inst  = '\\.byte',
  text_inst  = '\\.text',
  rtext_inst = '\\.rtext',
  rbyte_inst = '\\.rbyte',
  lbracket   = '\\(',
  rbracket   = '\\)',
  text       = '".*?"',
  comma      = ",",
  colon      = ":",
  equals     = '=',
  comment    = '(;[^\n]*)',
  x          = '(x|X)',
  y          = '(y|Y)',
  symbol     = '#?<?>?\\{.*?\\}',  # a symbol with evaluation
  symbol     = '[^\\s:,)]+'
)
```




Each line is split into tokens
------------------------------------------------------------------------------

* Each line is split into tokens
* any `comma`, `whitespace` or `comment` tokens are discarded
* any blank lines are discarded

```{r}
line_tokens <- c64asm::create_line_tokens(asm)
line_tokens
```


The main `prg_df` data structure is created from `line_tokens`
------------------------------------------------------------------------------

* All the computation/passes necessary to turn the ASM into actual bytes for PRG file
  take place as operations on this `prg_df` data.frame
* There are lots of columns here keeping track of values, cross-references to symbols and
  different representations of the values e.g. in both hexadecimal and decimal


```{r}
prg_df <- c64asm::create_prg_df(line_tokens)
```

```{r echo=FALSE}
prg_df %>% 
  select(-argbytes, -label_value, -op, -opcommand, -nargbytes, -nopbytes) %>%
  knitr::kable()
```


Calculate addresses and replace any symbols with these values
------------------------------------------------------------------------------

* symbols (such as `message`) may have their values defined by
    * their address, i.e. position in the bytestream relative to the start of the file
    * explicitly assigned a value e.g. `storage = $3000`
* Since the actual instruction address might only be specified at the very start of the file (using `* = $0820` for example), 
  then byte counting is used to figure out the address of all subsequent instructions.

```{r}
prg_df <- c64asm::process_symbols(prg_df)
```

```{r echo=FALSE}
prg_df %>% 
  select(-argbytes, -hexbytes, -label_value, -op, -symbol_op, -nopbytes, -nargbytes, -opcommand, -init_addr) %>% 
  select(-symbol_lo, -symbol_hi, -symbol_offset, -row_symbol_values) %>%
  knitr::kable()
```






Add zero padding so there are no gaps
------------------------------------------------------------------------------

* PRG files must represent a contiguous sequence of bytes - no gaps allowed!
* Find and insert sequences of zeros to ensure all gaps are filled.

```{r}
prg_df <- c64asm::process_zero_padding(prg_df)
```

```{r echo=FALSE}
prg_df %>% 
  select(-argbytes, -hexbytes, -label_value, -op, -symbol_op, -nopbytes, -nargbytes, -opcommand, -init_addr) %>% 
  select(-symbol_lo, -symbol_hi, -symbol_offset, -row_symbol_values) %>%
  knitr::kable()
```


Extract the bytes from `prg_df`
------------------------------------------------------------------------------

* `prg_df$bytes` is a list column which represents all the instruction bytes which 
  make up a PRG file.
* Use `purrr::flatten()` to convert to an integer vector (and remove any NULL entries)

```{r}
prg_df$bytes %>%
  purrr::flatten() %>%
  as.integer() %>% 
  as.raw()
```




Add the loading address to get the complete PRG
------------------------------------------------------------------------------

* PRG files are prefixed by their first address so they can be loaded in the 
  right location.
* Take the first address in `prg_df` and convert it to 2 bytes in low-byte/high-byte
  format using `c64asm::w2b()`


```{r}
# The following is equivalent to:  c64asm::extract_prg_bytes(prg_df)
as.raw(c(w2b(prg_df$addr[1]), as.integer(purrr::flatten(prg_df$bytes))))
```


Save the PRG
------------------------------------------------------------------------------

* Just write the bytes, with the loading address prefixed, 
directly to file with `writeBin()` - no other processing is needed.














